Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2023-06-01 paper/paperDraft.tex
literature remains

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}14{]}.

@@

Conceptually, the idea is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history {[}3{]},
  embeddings which do not protect certain classes may also be useful.}

@@

One of the first measures of bias has been developed in {[}1{]}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in {[}1{]}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(w\) is then understood as \(w\)'s projection
on the gender direction: \(\vec{w} \cdot gd\) (which, after normalizing
by dividing by \(\lVert w \rVert \,\lVert \mathsf{gd} \rVert\), is the
same as cosine similarity). Given a list \(N\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. We will have more to say about this assumption when we
  describe our dataset construction.} and the gender direction
\(\mathsf{gd}\), the direct gender bias is defined as the average cosine
similarity of the words in \(N\) from \(\mathsf{gd}\) (\(c\) is a
parameter determining how strict we want to be): \begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}5{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}16{]}.}

A measure of bias in word embeddings which does not employ similarities
to pre-calculated protected-class-level directions, the Word Embedding
Association Test (\textsf{WEAT}), has been proposed in {[}8{]}. The idea
here is that the bias between two sets of target words, \(X\) and \(Y\)
(we call them protected words), should be quantified in terms of the
cosine similarity between the protected words and attribute words coming
from two sets of stereotype attribute words, \(A\) and \(B\) (we will
call them attributes). For instance, \(X\) might be a set of male names,
\(Y\) a set of female names, \(A\) might contain stereotypically
male-related, and \(B\) stereotypically female-related career words. The
association difference for a particular word \(t\) (belonging to either
\(X\) or \(Y\)) is:

@@

\noindent The authors use it as a test statistic in some tests, and the
assumption that \(X\) and \(Y\) are of the same size is necessary for
the statistic to make sense. In the final measure (of effect size),
\textsf{WEAT}, means are taken, which makes it insensitive to size
differences between \(X\) and \(Y\): \begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}17{]} used in psychology, and it uses almost the same word sets,
allowing for a \emph{prima facie} sensible comparison with bias in
humans. In {[}8{]} the authors argue that significant biases---thus
measured--- similar to the ones discovered by IAT can be discovered in
word embeddings. In {[}11{]} the methodology is extended to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of cosine similarity does not make much difference,
while the bias effects vary greatly across embedding models
(interestingly, with social media-text trained embeddings being less
biased than those based on Wikipedia). A similar methodology is employed
in {[}4{]}. The authors employ word embeddings trained on corpora from
different decades to study the shifts in various biases.\footnote{Strictly
  speaking, these authors use Euclidean distances and their differences,
  but the way they take averages and averages thereof is analogous, and
  so what we will have to say about pre-averaging leading to false
  confidence applies to this methodology as well.}

@@

\textsf{WEAT}, however, has been developed to investigate biases
corresponding to a pair of supposedly opposing stereotypes, and so the
question arises as to how to generalize the measure to contexts in which
biases with respect to more than two stereotypical groups are to be
measured. Such a generalization can be found in {[}12{]}. The authors
introduce Mean Average Cosine distance (\textsf{MAC}) as a measure of
bias. Let \(T = \{t_1, \dots, t_k\}\) be a set of protected words, and
let each \(A_j\in \mathcal{A}\) be a set of attributes stereotypically
associated with a protected word where \(\mathcal{A}\). For instance,
when biases related to religion are to be investigated, they use a
dataset of the format illustrated in Table \ref{tab:religionOriginal}.
The measure is defined as follows:

@@

\begin{align*}
\mathsf{s}(t, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cosineDistance}(t,a) \\
\mathsf{mac}(T,\mathcal{A}) & = \frac{1}{\vert T \vert \,\vert \mathcal{A}\vert}\sum_{t \in T }\sum_{A_j \in \mathcal{A}}    \mathsf{s}(t,A_j)
\end{align*}

\noindent That is, for each protected word \(t\in T\), and each
attribute set \(A_j\), they first take the mean of distances for this
protected word and all attributes in a given attribute class, and then
take the mean of thus obtained means for all the protected words and all
the protected classes.\footnote{The authors' code are in their github
  repository,
  \url{https://github.com/TManzini/DebiasMulticlassWordEmbedding}.}

@@

Here is an example of results of debiasing that can be found in {[}12{]}
(Table \ref{tab:religionOriginal2}). It presents the results of applying
the procedure with respect to the Religion dataset. \textsf{MAC} scores
are given for the original datasets, and two resulting from two
debiasing methods applied to the Reddit embedding. The scores for
debiasing come with p-values for the comparison to the original
\textsf{MAC}.

@@

The first conceptual question we should ask is whether the initial
\textsf{MAC} values lower than 1 indeed are indicative of the presence
of bias? Of course, thinking abstractly, 1 is the ideal distance for
unrelated words. But clearly, there is variation in distances, which
might lead to non-biased lists also having \textsf{MAC} smaller than 1.
How much smaller? What may attract attention is the fact that the value
of cosine distance in ``Biased'' category is already quite high
(i.e.~close to 1) even before debiasing. High cosine distance indicates
low cosine similarity between values. One could think that average
cosine similarity equal to approximately 0.141 is not large enough to
consider it as bias. The authors, though, still aim to mitigate such
``bias'' to make the distance even larger. The question is, on what
basis is this small similarity still considered as a proof of the
presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group
of neutral attributes for comparison to obtain a more realistic gauge on
how to understand \textsf{MAC} values. Later on, in our approach, we
introduce such control word lists. One of them is a list of words we
intuitively considered neutral (see the Appendix for word lists).

Moreover, it might be the case that words that have to do with human
activities in general, even if unbiased, are systematically closer to
the protected words than merely neutral words. This, again, casts doubt
on whether comparing \textsf{MAC} to the abstractly ideal value of 1 is
a methodologically sound idea. For this reason we also use a second list
with intuitively non-stereotypical human attributes (again, see the
Appendix for a word list).

@@

In Figures (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the
empirical distributions, while paying attention to such divisions. The
horizontal lines represent the values of 1-\textsf{MAC} (that is, we now
talk in terms of cosine similarity rather than cosine distance; this is
a trivial linear transformation) that the authors considered indicative
of bias for stereotypes corresponding to given word lists. For instance,
in religion, \textsf{MAC} was .859 and considered a sign of bias, so we
plot \(0\pm (1-.859)\approx .14\) lines around similarity = 0 (that is,
distance = 1). Notice that most distributions are quite wide, and the
proportions of even neutral or human neutral words with similarities
higher than the value of 1-\textsf{MAC} deserving debiasing according to
the authors is quite high.

@@

The approaches we have been describing use means of mean average cosine
similarities to measure similarity between protected words and
attributes coming from harmful stereotypes. However, a closer inspection
look at the individual values that are taken for the calculations, it
turns out that there are quite a few outliers and surprisingly
dissimilar words. This problem becomes transparent when we examine the
visualizations of the individual cosine distances, following the idea
that one of the first steps in understanding data is to look at it.
Let's start with inspecting two examples in Figures \ref{fig:muslim} and
\ref{fig:priest}.

@@

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/visMuslimd-1} \end{center}
\caption{Actual distances for the protected word "muslim". The most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. }
\label{fig:muslim}
\end{figure}

@@

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/visPriestd-1} \end{center}

\caption{Actual distances for the protected word "priest", the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.}
\label{fig:priest}
\end{figure}

With such a method the uncertainty involved is not really considered
which makes it even more difficult to give reasonable interpretations of
the results. We propose the use of Bayesian method to obtain some
understanding of the influence the uncertainty has on the interpretation
of final results.

\noindent \(s(X,Y,A,B)\) is the statistic used in the significance test,
and the \(p\)-value is obtained by bootstrapping: it is the frequency of
\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
the difference in means as follows:

\footnotesize

\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

@@

The t-tests they employ are run on average cosines used to calculate
\textsf{MAC}.

The effect sizes reported by {[}8{]} for lists of words for which the
embeddings are supposedly biased range from 2.06 to 1.81, and the
reported \(p\)-values are in the range of \(10^{-7}-10^{-2}\) with one
exception for Math vs arts, where it is \(.018\). The question is, are
those results meaningful?

One way to answer this question is to think in terms of null models. If
the words actually are samples from two populations with equal mean, how
often would we see effect sizes in this range? How often would we reach
the \(p\)-values that the authors reported?

We start our exploration with considering a list of 16 protected words,
each class containing 8, with 16 attributes, 8 in each attribute class.
In each particular sample we will investigate the consequence of the
model being in fact null: we draw the cosine distances, every time, from
the \(\mathsf{Normal}(0,.08)\) distribution. \(0.08\) is approximately
the empirical standard deviation observed in fairly large samples of
neutral words, and 16 is the sample size used in the WEAT7 word list,
which is not much different from the other sample sizes in word lists
used by {[}8{]}

So let's draw one sample of this type. We calculate the s-values, the
test statistics and the effect sizes. Then, following the original
methodology, we obtain bootstrapped distributions of the effect sizes
and the test statistics by calculating the same for each possible equal
split of the initial random data set. The observed test statistic is
0.386222 and 1.2671948. The bootstrapped distributions of the test
statistics and effect sizes are illustrated in Figure
\ref{fig:caliskanCalc}. Quite notably both (two-sided) \(p\) values are
uneven and rather low.

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:caliskanCalcB2B-1} \end{center}

\caption{Bootstrapped distributions of test statistics and effect sizes in a random sample given the null hypothesis.}
\label{fig:caliskanCalc}
\end{figure}

At this point, we might think---right, so while drawing on the
assumption of the null hypothesis we just stumbled into a data set that
randomly happened to display strong bias. We decide to double-check this
by visual inspection expecting exactly this: a strong, clearly visible
bias (Figure \ref{fig:caliskanDistances}).

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:caliskanDistC2B-1} \end{center}

@@

In the calculations of \textsf{WEAT} means are taken twice. The
\(s\)-values themselves are means, and then means of \(s\)-values are
compared between groups. The test statistic used itself is of this sort
as well---while it compares sums, these are sums of sets of the same
size, so they are really linear transformations of means of means.
Statistical troubles start when we run statistical tests on sets of
means, for at least two reasons One, by pre-averaging data we throw away
information about sample sizes. For the former point, think about
proportions: 10 out of 20 and 2 out of 4 give the same mean, but you
would obtain more information by making the former observation rather
than by making the latter. And especially in this context, in which the
word lists are not huge, sample sizes should matter. Two, when we
pre-average, we remove variation, and so pre-averaging tend to
manufacture false confidence. Group means display less variation than
the raw data points, and the standard deviation of a set of means of
sets of means is bound to be lower than the original standard deviation
in the row data. Now, if you calculate your effect size by dividing by
the pre-averaged standard deviation, your quite likely to get something
that looks like a strong effect size, but the results of your
calculations might not track anything interesting.

So let us think again about the question that we are ultimately
interested in. Are the \(X\) terms systematically closer to (further
from) the \(A\) attributes (\(B\) attributes) than the \(Y\) words? But
now let's use the raw data points to try to answer these questions. As
the first stab, let us run two quick \(t\)-tests to gauge what the raw
data illustrated in Figure \ref{fig:caliskanDistances} tell us.

First, distances to \(A\) attributes for \(X\) words and \(Y\) words.
Whoa, the result is significant! The \(p\)-value is \(0.02\). So sure,
the sample is in some sense unusual. But the 95\% confidence interval
for the difference in means is \([.0052 .061]\), clearly nothing that a
reader would expect given that the calculated effect size seemed quite
large. How about the \(B\) attributes? Here the \(p\)-value is \(.22\)
and the 95\% confidence interval is \([-0.03, .009]\), even less of a
reason to think a bias is present.

The difficulties are exacerbated by the fact that statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way {[}8{]} did. Do
this 10000 times and look at what the distributions of these values are
on the assumption of the null model with realistic empirically motivated
raw data point standard deviation.

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:ourDistancesPlot2e-1} \end{center}

\caption{Distributions of test statistics and effect sizes based on 10k simulations on the assumption of a null model in which all distances come from normal distribution with $\mu =0, \sigma = .08, n=10k$.}
\label{fig:ourDistances}
\end{figure}

>>>>

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}15{]}.

@@

The underlying intuition is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history {[}3{]},
  embeddings which do not protect certain classes may also be useful.}

@@

One of the first measures of bias has been developed in {[}1{]}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in {[}1{]}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(w\) is then understood as \(w\)'s projection
on the gender direction: \(\vec{w} \cdot gd\) (which, after normalizing
by dividing by \(\lVert w \rVert \,\lVert \mathsf{gd} \rVert\), is the
same as cosine similarity). Given a list \(N\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. See our list in Appendix \ref{app:custom} and a brief
  discussion in Subsection \ref{subsec:interpretability}.} and the
gender direction \(\mathsf{gd}\), the direct gender bias is defined as
the average cosine similarity of the words in \(N\) from \(\mathsf{gd}\)
(\(c\) is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}5{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}17{]}.}

A measure of bias in word embeddings which does not proceed by
identifying bias directions (such as a gender vector), the Word
Embedding Association Test (\textsf{WEAT}), has been proposed in
{[}8{]}. The idea here is that the bias between two sets of target
words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we will call them attributes). For instance, \(X\)
might be a set of male names, \(Y\) a set of female names, \(A\) might
contain stereotypically male-related, and \(B\) stereotypically
female-related career words. The association difference for a particular
word \(t\) (belonging to either \(X\) or \(Y\)) is:

@@

\noindent The intention is that large values of \(s\) scores suggest
systematic differences between how \(X\) and \(Y\) are related to \(A\)
and \(B\), and therefore are indicative of the presence of bias. The
authors use it as a test statistic in some tests,\footnote{Note their
  method assumes \(X\) and \(Y\) are of the same size.} and the final
measure of effect size, \textsf{WEAT}, is constructed by taking means of
these values and standardizing: \begin{align} \label{eq:weat}
\mathsf{WEAT}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}18{]} used in psychology, and in some applications it uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. In {[}8{]} the authors argue that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. In {[}11{]} the methodology is extended
to a multilingual and cross-lingual setting, arguing that using
Euclidean distance instead of cosine similarity does not make much
difference, while the bias effects vary greatly across embedding
models.\footnote{Interestingly, with social media-text trained
  embeddings being less biased than those based on Wikipedia.} A similar
methodology is employed in {[}4{]}. The authors employ word embeddings
trained on corpora from different decades to study the shifts in various
biases through the century.\footnote{Strictly speaking, these authors
  use Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

@@

\textsf{WEAT} has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, so the question arises as to
how to generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in {[}12{]}. The authors introduce Mean
Average Cosine distance (\textsf{MAC}) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\begin{align*}
\mathsf{s}(t, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cosineDistance}(t,a) \\
\mathsf{MAC}(T,\mathcal{A}) & = \frac{1}{\vert T \vert \,\vert \mathcal{A}\vert}\sum_{t \in T }\sum_{A_j \in \mathcal{A}}    \mathsf{s}(t,A_j)
\end{align*}

\noindent That is, for each protected word \(t\in T\), and each
attribute set \(A_j\), they first take the mean of distances for this
protected word and all attributes in a given attribute class, and then
take the mean of thus obtained means for all the protected words and all
the protected classes.\footnote{The authors' code is available through
  their github repository at
  \url{https://github.com/TManzini/DebiasMulticlassWordEmbedding}.}

@@

Table \ref{tab:religionOriginal2} contains an example of \textsf{MAC}
scores (and \(p\) values, we explain how these are obtained in
Subsection \ref{subsec:problems}) before and after deploying two
debiasing methods to the Reddit embedding, where the score is calculated
using the Religion word lists from {[}12{]}. For our purpose the details
of the debiasing method are not important: what matters is that
\textsf{MAC} is used in the evaluation of these methods.

@@

The first question we should ask is whether the initial \textsf{MAC}
values lower than 1 indeed are indicative of the presence of bias?
Thinking abstractly, 1 is the ideal distance for unrelated words. But in
fact there is some variation in distances, which might lead to
non-biased lists also having \textsf{MAC} scores smaller than 1. How
much smaller? What may attract attention is the fact that the value of
cosine distance in ``Biased'' category is already quite high (i.e.~close
to 1) even before debiasing. High cosine distance indicates low cosine
similarity between values. One could think that average cosine
similarity equal to approximately 0.141 is not large enough to claim the
presence of a bias to start with. The authors, though, still aim to
mitigate it by making the distances involved in the \textsf{MAC}
calculations even larger. The question is, on what basis is this small
similarity still considered as a proof of the presence of bias, and
whether these small changes are meaningful.

The problem is that the original paper did not employ any control group
of neutral attributes for comparison to obtain a more realistic gauge on
how to understand \textsf{MAC} values. Later on, in our approach, we
introduce such control word lists. One of them is a list of words we
intuitively considered neutral. Moreover, it might be the case that
words that have to do with human activities in general, even if
unbiased, are systematically closer to the protected words than merely
neutral words. This, again, casts doubt on whether comparing
\textsf{MAC} to the abstractly ideal value of 1 is a methodologically
sound idea. For this reason we also use a second list with intuitively
non-stereotypical human attributes.\footnote{See Appendix
  \ref{app:custom} for the word lists.}

@@

In Figures (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the
empirical distributions, while paying attention to such divisions. The
horizontal lines represent the values of \(1 - \mathsf{MAC}\) (that is,
we now talk in terms of cosine similarity rather than cosine distance)
that the authors considered indicative of bias for stereotypes
corresponding to given word lists. For instance, in religion,
\textsf{MAC} was .859, which was considered a sign of bias, so we plot
\(0\pm (1-.859)\approx .14\) lines around similarity = 0 (that is,
distance = 1). Notice that most distributions are quite wide, and the
proportions of even neutral or human neutral words with similarities
higher than the value of \(1 - \mathsf{MAC}\) deserving debiasing
according to the authors are quite high.

@@

The approaches we have been describing use means of mean average cosine
similarities to measure similarity between protected words and
attributes coming from harmful stereotypes. But once we take a look at
the individual values, it turns out that the raw data variance is rather
high, and there are quite a few outliers and surprisingly dissimilar
words. This problem becomes transparent when we examine the
visualizations of the individual cosine distances, following the idea
that one of the first steps in understanding data is to look at it.
Let's start with inspecting two examples of such visualizations in
Figures \ref{fig:muslim} and \ref{fig:priest} (we also include neutral
and human predicates to make our point more transparent). Again, we
emphasize that \textbf{we do not condone the associations which we are
about to illustrate.}

@@

\begin{center}\includegraphics[width=0.9\linewidth]{paperDraft_files/figure-latex/visMuslimd-1} \end{center}
\caption{Actual distances for the protected word \textsf{muslim}. }
\label{fig:muslim}
\end{figure}

@@

\begin{center}\includegraphics[width=0.9\linewidth]{paperDraft_files/figure-latex/visPriestd-1} \end{center}

\caption{Actual distances for the protected word \textsf{priest}.}
\label{fig:priest}
\end{figure}

As transparent in Figures \ref{fig:muslim} and \ref{fig:priest}, for the
protected word \textsf{muslim}, the most similar attributes tend to be
the ones associated with it stereotypically, but then words associated
with other stereotypes come closer than neutral or human predicates. For
the protected word \textsf{priest}, the situation is even less as
expected: the nearest attributes are human attributes, and all there
seems to be no clear pattern to the distances to other attributes.

The general phenomenon that makes us skeptical about running statistical
tests on pre-averaged data is that raw datasets of different variance
can result in the same pre-averaged data and consequently the same
single-number metric. In other words, a method that proceeds this way is
not very sensitive to the real sample variance.

Let us illustrate how this problem arises in the context of
\textsf{WEAT}. Once a particular \(s(X,Y,A,B)\) is calculated, the
question arises as to whether a value that high could have arisen by
chance. To address the question, each \(s(X,Y,A,B)\) is used in the
original paper to generate a \(p\)-value by bootstrapping. The
\(p\)-value is the frequency of how often it is the case that
\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for sampled equally sized partitions
\(X_i, Y_i\) of \(X\cup Y\). The WEAT score is then computed by
standardizing the difference in means of means by dividing by the
standard deviation of means, see equation \eqref{eq:weat}.

@@

\noindent The \textsf{WEAT} scores reported by {[}8{]} for lists of
words for which the embeddings are supposedly biased range from 2.06 to
1.81, and the reported \(p\)-values are in the range of
\(10^{-7}-10^{-2}\) with one exception for \emph{Math vs Arts}, where it
is \(.018\).

The question is, are those results meaningful? One way to answer this
question is to think in terms of null generative models. If the words
actually are samples from two populations with equal means, how often
would we see \textsf{WEAT} scores in this range? How often would we
reach the \(p\)-values that the authors reported?

Imagine there are two groups of protected words, each of size 8, and two
groups of stereotypical attributes, of the same size.\footnote{16 is the
  sample size used in the WEAT7 word list, which is not much different
  from the other sample sizes in word lists used by {[}8{]}).} Each such
a collection of samples, as far as our question is involved, is
equivalent to a sample of \(16^2\) cosine distances. Further, imagine
there really is no difference between these groups of word and the model
is in fact null. That is, we draw the cosine distances from the
\(\mathsf{Normal}(0,.08)\) distribution.\footnote{\(.08\) is
  approximately the empirical standard deviation observed in fairly
  large samples of neutral words.}

In Figure \ref{fig:caliskanCalc} we illustrate one iteration of the
procedure. We draw one such sample of size \(16^2\). Then we actually
list all possible ways to split the 16 words in two equal sets (each
such a split is one bootstrapped sample) and for each of them we
calculate the \textsf{s} values and \textsf{WEAT}. What are the
resulting distributions of \(\textsf{s}\) scores and what \(p\)-values
do they lead to? What are the resulting effect sizes for each
bootstrapped sample, and how often can we get an effect size as large as
the ones reported in the original paper?

In the bootstrapped samples we would rather expect low \textsf{s} values
and low \textsf{WEAT}: after all, these are just random permutations of
random distances all of which are drawn from the same null distribution.
Let's take a look at one such a bootstrapped sample.

On purpose we picked a rather unusual one: the observed test statistic
is 0.39 and 1.27. The bootstrapped distributions of the test statistics
and effect sizes are illustrated in Figure \ref{fig:caliskanCalc},
together with this particular example. Quite notably both (two-sided)
\(p\) values for our example are rather low (Figure
\ref{fig:caliskanCalc}). These facts might suggested that we ended up
with a situation where ``bias'' is present (albeit, due to random
noise). The reason why we picked it is that it is an example of a word
list that ends up with relatively low \(p\)-value and a relatively
unusual effect size, but nevertheless its closer inspection shows that
even for a word list with such properties there is no clear reason to
think that the bias is present.

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:caliskanCalcB2D-1} \end{center}

\caption{Bootstrapped distributions of test statistics and effect sizes in a random sample given the null hypothesis. We used a sample from the null  model with N(0,.08) and 16 protected words, and then boostrapped from it, following the original methodology. One particular boostrapped sample is highlighted, and discussed further in the text. }
\label{fig:caliskanCalc}
\end{figure}

At this point, we might think that we just stumbled into a bootstrapped
sample that randomly happened to display strong bias. We decide to
double-check this by visual inspection expecting exactly this: a strong,
clearly visible bias (Figure \ref{fig:caliskanDistances}).

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:caliskanDistC2K-1} \end{center}

@@

In the calculations of \textsf{WEAT} means are taken twice. The
\textsf{s}-values themselves are means, and then means of
\textsf{s}-values are compared between groups. Statistical troubles
start when we run statistical tests on sets of means, for at least two
reasons.

\begin{enumerate}

\item By pre-averaging data we throw away information about sample sizes. For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.  And especially in this context, in which the word lists are not huge, sample sizes should matter.

\item When we pre-average, we disregard variation, and therefore pre-averaging  tend to manufacture false confidence. Group means display less variation than the raw data points, and the standard deviation of a set of means of sets of means is bound to be lower than the original standard deviation in the row data. Now, if you calculate your effect size by dividing by the pre-averaged standard deviation, your quite likely to get something that looks like a strong effect size, but the results of your calculations might not track anything interesting.
\end{enumerate}

Let us think again about the question that we are ultimately interested
in. Are the \(X\) terms systematically closer to (further from) the
\(A\) attributes (\(B\) attributes) than the \(Y\) words? But now let's
use the raw data points to try to answer these questions.

To start with, let us run two quick \(t\)-tests to gauge what the raw
data illustrated in Figure \ref{fig:caliskanDistances} tell us. First,
distances to \(A\) attributes for \(X\) words and \(Y\) words. Well, the
result is---strictly speaking---statistically significant. The
\(p\)-value is \(0.02\) (more than ten times higher than the
\(p\)-valued obtained by the boostratpping procedure. So the sample is
in some sense unusual. But the 95\% confidence interval for the
difference in means is \([.0052, .061]\), clearly nothing that a reader
would expect given that the calculated effect size seemed quite large.
How about the distances to the \(B\) attributes? Here the \(p\)-value is
\(.22\) and the 95\% confidence interval is \([-0.03, .009]\), even less
of a reason to think a bias is present.

The difficulties are exacerbated by the fact that statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way {[}8{]} did. Do
this 10000 times, each time calculating \textsf{WEAT} and \textsf{s}
values, and look at what the distributions of these values are on the
assumption of the null model with realistic empirically motivated raw
data point standard deviation.

@@

\begin{center}\includegraphics[width=0.8\linewidth]{paperDraft_files/figure-latex/fig:ourDistancesPlot2f-1} \end{center}

\caption{Distributions of test statistics and effect sizes based on 10k simulations on the assumption of a null model in which all distances come from normal distribution with $\mu =0, \sigma = .08, n=10k$. We also mark the sample we have been using as an example.}
\label{fig:ourDistances}
\end{figure}
